A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-17 tidyindex.qmd
go through the drought example

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#options(width = 70)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(tidyindex)
library(lmomco)
library(lubridate)
library(SPEI)
library(tsibble)
library(GGally)
library(patchwork)
library(tourr)
```

@@

The rest of the paper is structured as follows: @sec-tidy reviews the tidy framework in R and how index construction can benefit from such a framework. The details of the pipeline modules are presented in @sec-pipeline. @sec-dev explains the design of the `tidyindex` pacakge that implements the modules. Examples are given in @sec-examples to illustrate the use cases of the pipeline.

@@

# Software design {#sec-dev}

@@

A common task for drought researchers is to compute indexes at different parameter combinations. This can be used to identify the spatial and temporal extent of drought events, recommend the best parameter choice, or compare the effectiveness of indexes for monitoring drought. The example below computes two indexes: SPI and SPEI, at various time scales and fitted distributions, for stations in the state of Queensland in Australia. The purpose of the example is to demonstrate the interfaces the tidyindex package built to allow easy computing at different parameter combinations.

The state of Queensland in Australia is frequently affected by natural disaster events such as flood and drought, which can have significant impacts on its agricultural industry. This study uses daily data from Global Historical Climatology Network Daily (GHCND), accessed via the package `rnoaa` to examine drought/flood condition in Queensland. Daily data is average into monthly and stations are excluded if monthly data contains missings, which is required for calculating both SPI and SPEI. This gives `r length(unique(queensland$id))` stations with complete records from 1990 January to 2022 April.

```{r fig-pipeline-spei}
#| fig-align: center
#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-spei.png"))
```

The function `compute_indexes()` can be used to collectively compute multiple indexes. The tidyindex offers wrapper functions, with the prefix idx\_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function. For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise). Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs. The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)). The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.

@@

The output from `compute_indexes()` contains index values and associated parameter in a long tibble. It includes the original variables (`id`, `ym`, `prcp`, `tmax`, `tmin`, `tavg`, `long`, `lat`, and `name`), index parameters (`.idx`, `.scale`, `.method`, and `.dist`), intermediate variables (`pet`, `.agg`, and `.fitted`), and the final index (`.index`). This data can be visualised across space or time, or simultaneously, to explore the wet/dry condition in Queensland. @fig-compute-spatial visualises the spatial distribution of SPI at two periods (2010 October - 2011 March and 2019 October - 2020 March) with significant natural disaster events: 2010/11 Queensland flood and 2019 Australia drought, which contributes to the notorious 2019/20 bushfire. @fig-compute-temporal displays the sensitivity of the SPEI series for one particular station, Texas post office, at different time scales and fitted distributions. These two plots demonstrate some possibilities to explore the indexes after they are computed from `compute_indexes()`.

```{r fig-compute-spatial}
#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20. The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
#| fig-width: 6
load(here::here("data/idx.rda"))
queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
  rmapshaper::ms_simplify(keep = 0.02)
queensland_map %>%
  ggplot() +
  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
  geom_point(
    data = idx %>%
      filter(.idx == "spi", .scale == 12) %>%
      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )), 
    aes(x = long, y = lat, color = .index)) +
  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
  theme_void() +
  facet_wrap(vars(ym), ncol = 6)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(width = 70)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(tidyindex)
library(lmomco)
library(lubridate)
library(SPEI)
library(tsibble)
library(GGally)
library(patchwork)
library(tourr)
```

@@

The rest of the paper is structured as follows: @sec-tidy reviews the tidy framework in R and how index construction can benefit from such a framework. The details of the pipeline modules are presented in @sec-pipeline. @sec-software explains the design of the `tidyindex` pacakge that implements the modules. Examples are given in @sec-examples to illustrate the use cases of the pipeline.

@@

# Software design {#sec-software}

@@

The state of Queensland in Australia frequently experience natural disaster events such as flood and drought, which can significantly impact its agricultural industry. This example uses daily data from Global Historical Climatology Network Daily (GHCND), aggregated into monthly precipitation, to compute two drought indexes -- SPI and SPEI -- at various time scales and fitted distributions, for  `r length(unique(queensland$id))` stations in the state of Queensland in Australia, spanning from January 1990 to April 2022. This example showcases the basic calculation of indexes with different parameter specifications within the pipeline framework. The dataset used in this example is available in the `tidyindex` package as `queensland` and blow prints the first few rows of the data:

```{r}
queensland %>% head(5)
```

```{r fig-spei}
#| fig-align: center
#| fig-cap: "Index pipeline for two drought indexes: the Standardized Precipitation Index (SPI) and the Standardized Precipitation-Evapotranspiration Index (SPEI). Both indexes share similar construction steps with SPEI having two steps additional steps (variable transformation and dimension reduction) to convert temperature into evapotranspiration and combine it with the precipitation series. "
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(path = here::here("figures/pipeline-spei.png"))
```

@fig-spei illustrates the pipeline steps of the two indexes. The two indexes are similar with the distinct that SPEI involves two additional steps -- variable transformation and dimension reduction -- prior to temporal processing. As introduced in @sec-software, wrapper functions are available for both indexes as `idx_spi()` and `idx_spei()`, which allows for the specification of different time scales and distributions for fitting the aggregated series. In `tidyindex`, multiple indexes can be calculated collectively using the function `compute_indexes()`. Both SPI and SPEI are calculated across four time scales (6, 12, 24, and 36 months). The SPEI is fitted with two distributions (log-logistic and general extreme value distribution) and the gamma distribution is used for SPI:

@@

```{r}
load(here::here("data/idx.rda"))
```

```{r}
idx
```

The output contains the original data, index values (`.index`), parameter used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`pet`, `.agg`, and `.fitted`). This data can be visualized to investigate the spatio-temporal distribution of the drought/flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations. @fig-compute-spatial and @fig-compute-temporal exemplify two possibilities. @fig-compute-spatial presents the spatial distribution of SPI during two periods: October 2010 to March 2011 for the 2010/11 Queensland flood and October 2019 to March 2020 for the 2019 Australia drought, which contributes to the notorious 2019/20 bushfire. @fig-compute-temporal displays the sensitivity of the SPEI series at the Texas post office to different time scales and fitted distributions. Larger time scales produce a smoother index across time, however, all time scales indicate an extreme drought (corresponding to -2 in SPEI) in 2020, confirming the severity of the drought across different time horizons. Moreover, the chosen distribution has less influence on the index, with general extreme value distribution tending to produce more extreme outcomes than log logistic distribution for the extreme events (index > 2 or <-2).

```{r fig-compute-spatial}
#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20. The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
#| fig-width: 6
queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
  rmapshaper::ms_simplify(keep = 0.02)
queensland_map %>%
  ggplot() +
  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
  geom_point(
    data = idx %>%
      filter(.idx == "spi", .scale == 12) %>%
      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )),
    aes(x = long, y = lat, color = .index)) +
  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
  theme_void() +
  facet_wrap(vars(ym), ncol = 6)
```

### 2023-08-17 tidyindex.tex
go through the drought example

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While many
work has focused on defining new indexes for the application of
interest, less attention has been directed towards understanding how the
a given index behave under different scenarios. This research introduces
a data pipeline comparising 9 modules to assemble indexes from a series
of modular building blocks. This modular data pipeline is universally
applicable to index computation and allows analysts to incorporate the
analysis on index into the construction pipeline. Analysts can compute
with different index parameters, adjust steps in the index definition by
adding, removing, and swapping them to experiment with various index
designs, calculate uncertainty measures, and assess index robustness .
The paper presents two examples to illustrate the usage of the pipeline.
The first example claculates two drought indexes with different
parameters to investigate the spatio-temporal distribution of drought in
Queensland, Australia. The second example on the Global Gender Gap Index
(GGGI) explores the effects of adjusting linear combination weights on a
country's index values and rankings.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

The rest of the paper is structured as follows: Section~\ref{sec-tidy}
reviews the tidy framework in R and how index construction can benefit
from such a framework. The details of the pipeline modules are presented
in Section~\ref{sec-pipeline}. Section~\ref{sec-dev} explains the design
of the \texttt{tidyindex} pacakge that implements the modules. Examples
are given in Section~\ref{sec-examples} to illustrate the use cases of
the pipeline.

@@

\hypertarget{sec-dev}{%
\section{Software design}\label{sec-dev}}

@@

A common task for drought researchers is to compute indexes at different
parameter combinations. This can be used to identify the spatial and
temporal extent of drought events, recommend the best parameter choice,
or compare the effectiveness of indexes for monitoring drought. The
example below computes two indexes: SPI and SPEI, at various time scales
and fitted distributions, for stations in the state of Queensland in
Australia. The purpose of the example is to demonstrate the interfaces
the tidyindex package built to allow easy computing at different
parameter combinations.

The state of Queensland in Australia is frequently affected by natural
disaster events such as flood and drought, which can have significant
impacts on its agricultural industry. This study uses daily data from
Global Historical Climatology Network Daily (GHCND), accessed via the
package \texttt{rnoaa} to examine drought/flood condition in Queensland.
Daily data is average into monthly and stations are excluded if monthly
data contains missings, which is required for calculating both SPI and
SPEI. This gives 29 stations with complete records from 1990 January to
2022 April.

@@

\caption{\label{fig-pipeline-spei}Diagram of pipeline steps for index
construction. will need to be updated with better design and the
distribution fitting step.}

@@

The function \texttt{compute\_indexes()} can be used to collectively
compute multiple indexes. The tidyindex offers wrapper functions, with
the prefix idx\_, that simplify the calculation of commonly used indexes
by combining a set of pipeline steps into a single function. For
example, the function \texttt{idx\_spei()} includes the five steps
previously described in \textbf{?@sec-toy-example} (variable
transformation, dimension reduction, temporal aggregation, distribution
fit, and normalise). Each \texttt{idx\_xxx()} function specifies the
relevant parameters relevant to the index: the thornthwaite method is
used to calculate PET in SPEI, with the average temperature
(\texttt{tavg}) and latitude (\texttt{lat}) used as inputs. The SPEI is
computed at four time scales (6, 12, 24, and 36 months) and fitted with
two distributions (Log-logistic and General Extreme Value (GEV)). The
SPI is also computed at the same four time scales and uses the default
gamma distribution to fit the aggregated series.

@@

The output from \texttt{compute\_indexes()} contains index values and
associated parameter in a long tibble. It includes the original
variables (\texttt{id}, \texttt{ym}, \texttt{prcp}, \texttt{tmax},
\texttt{tmin}, \texttt{tavg}, \texttt{long}, \texttt{lat}, and
\texttt{name}), index parameters (\texttt{.idx}, \texttt{.scale},
\texttt{.method}, and \texttt{.dist}), intermediate variables
(\texttt{pet}, \texttt{.agg}, and \texttt{.fitted}), and the final index
(\texttt{.index}). This data can be visualised across space or time, or
simultaneously, to explore the wet/dry condition in Queensland.
Figure~\ref{fig-compute-spatial} visualises the spatial distribution of
SPI at two periods (2010 October - 2011 March and 2019 October - 2020
March) with significant natural disaster events: 2010/11 Queensland
flood and 2019 Australia drought, which contributes to the notorious
2019/20 bushfire. Figure~\ref{fig-compute-temporal} displays the
sensitivity of the SPEI series for one particular station, Texas post
office, at different time scales and fitted distributions. These two
plots demonstrate some possibilities to explore the indexes after they
are computed from \texttt{compute\_indexes()}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While many
work has focused on defining new indexes for the application of
interest, less attention has been directed towards understanding how the
a given index behave under different scenarios. This research introduces
a data pipeline comparising 9 modules to assemble indexes from a series
of modular building blocks. This modular data pipeline is universally
applicable to index computation and allows analysts to incorporate the
analysis on index into the construction pipeline. Analysts can compute
with different index parameters, adjust steps in the index definition by
adding, removing, and swapping them to experiment with various index
designs, calculate uncertainty measures, and assess index robustness .
The paper presents two examples to illustrate the usage of the pipeline.
The first example claculates two drought indexes with different
parameters to investigate the spatio-temporal distribution of drought in
Queensland, Australia. The second example on the Global Gender Gap Index
(GGGI) explores the effects of adjusting linear combination weights on a
country's index values and rankings.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

The rest of the paper is structured as follows: Section~\ref{sec-tidy}
reviews the tidy framework in R and how index construction can benefit
from such a framework. The details of the pipeline modules are presented
in Section~\ref{sec-pipeline}. Section~\ref{sec-software} explains the
design of the \texttt{tidyindex} pacakge that implements the modules.
Examples are given in Section~\ref{sec-examples} to illustrate the use
cases of the pipeline.

@@

\hypertarget{sec-software}{%
\section{Software design}\label{sec-software}}

@@

The state of Queensland in Australia frequently experience natural
disaster events such as flood and drought, which can significantly
impact its agricultural industry. This example uses daily data from
Global Historical Climatology Network Daily (GHCND), aggregated into
monthly precipitation, to compute two drought indexes -- SPI and SPEI --
at various time scales and fitted distributions, for 29 stations in the
state of Queensland in Australia, spanning from January 1990 to April
2022. This example showcases the basic calculation of indexes with
different parameter specifications within the pipeline framework. The
dataset used in this example is available in the \texttt{tidyindex}
package as \texttt{queensland} and blow prints the first few rows of the
data:

\begin{verbatim}
# A tibble: 5 x 9
  id                ym  prcp  tmax  tmin  tavg  long   lat name       
  <chr>          <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <chr>      
1 ASN00029038 1990 Jan  1682  34.3  24.7  29.5  142. -15.5 KOWANYAMA ~
2 ASN00029038 1990 Feb   416  35.2  23.2  29.2  142. -15.5 KOWANYAMA ~
3 ASN00029038 1990 Mar  2026  32.5  23.6  28.0  142. -15.5 KOWANYAMA ~
4 ASN00029038 1990 Apr   597  32.9  17.7  25.3  142. -15.5 KOWANYAMA ~
5 ASN00029038 1990 May   244  31.8  20.1  25.9  142. -15.5 KOWANYAMA ~
\end{verbatim}

@@

\caption{\label{fig-spei}Index pipeline for two drought indexes: the
Standardized Precipitation Index (SPI) and the Standardized
Precipitation-Evapotranspiration Index (SPEI). Both indexes share
similar construction steps with SPEI having two steps additional steps
(variable transformation and dimension reduction) to convert temperature
into evapotranspiration and combine it with the precipitation series.}

@@

Figure~\ref{fig-spei} illustrates the pipeline steps of the two indexes.
The two indexes are similar with the distinct that SPEI involves two
additional steps -- variable transformation and dimension reduction --
prior to temporal processing. As introduced in
Section~\ref{sec-software}, wrapper functions are available for both
indexes as \texttt{idx\_spi()} and \texttt{idx\_spei()}, which allows
for the specification of different time scales and distributions for
fitting the aggregated series. In \texttt{tidyindex}, multiple indexes
can be calculated collectively using the function
\texttt{compute\_indexes()}. Both SPI and SPEI are calculated across
four time scales (6, 12, 24, and 36 months). The SPEI is fitted with two
distributions (log-logistic and general extreme value distribution) and
the gamma distribution is used for SPI:

@@

\begin{verbatim}
# A tibble: 128,586 x 19
   .idx  .period id             ym  prcp  tmax  tmin  tavg  long   lat
   <chr>   <dbl> <chr>       <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
 1 spei        6 ASN0002~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
 2 spei        6 ASN0002~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
 3 spei        6 ASN0002~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7
 4 spei        6 ASN0002~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7
 5 spei        6 ASN0003~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9
 6 spei        6 ASN0003~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9
 7 spei        6 ASN0003~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1
 8 spei        6 ASN0003~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1
 9 spei        6 ASN0003~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6
10 spei        6 ASN0003~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6
# i 128,576 more rows
# i 9 more variables: name <chr>, pet <dbl>, diff <dbl>,
#   .scale <dbl>, .agg <dbl>, .method <chr>, .fitted <dbl>,
#   .dist <chr>, .index <dbl>
\end{verbatim}

The output contains the original data, index values (\texttt{.index}),
parameter used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
and all the intermediate variables (\texttt{pet}, \texttt{.agg}, and
\texttt{.fitted}). This data can be visualized to investigate the
spatio-temporal distribution of the drought/flood events, as well as the
response of index values to different time scales and distribution
parameters at specific single locations.
Figure~\ref{fig-compute-spatial} and Figure~\ref{fig-compute-temporal}
exemplify two possibilities. Figure~\ref{fig-compute-spatial} presents
the spatial distribution of SPI during two periods: October 2010 to
March 2011 for the 2010/11 Queensland flood and October 2019 to March
2020 for the 2019 Australia drought, which contributes to the notorious
2019/20 bushfire. Figure~\ref{fig-compute-temporal} displays the
sensitivity of the SPEI series at the Texas post office to different
time scales and fitted distributions. Larger time scales produce a
smoother index across time, however, all time scales indicate an extreme
drought (corresponding to -2 in SPEI) in 2020, confirming the severity
of the drought across different time horizons. Moreover, the chosen
distribution has less influence on the index, with general extreme value
distribution tending to produce more extreme outcomes than log logistic
distribution for the extreme events (index \textgreater{} 2 or
\textless-2).
